Setting parameters for biological models with ANIMO

### 2014-03-07 Paper.tex
Updating according to reviews (added more references)

A basic type of analysis that can be performed with ANIMO is the generation of
a simulation run, which is presented to the user in the form of a graph (to the right). The activity
graphs generated by ANIMO show the variation in activity of selected nodes over the course of the simulation run.
In addition to that, a slider placed under each graph allows the user to color the nodes in the Network panel
depending on their activity level at any point during the course of the simulation: the Legend
panel on the left links colors to activity. ANIMO models are based on the formalism of \tas~\cite{timed-automata-alur-dill},
which is not directly shown to the user. Indeed, ANIMO was designed with the aim of enabling expert biologists
to formalize their knowledge without the need for additional mathematical training.

\begin{figure}[htbp]
\begin{center}
\subfloat[\label{fig:animo-ui}]{\includegraphics[width=.55\textwidth]{images/animo-ui}} \quad
\subfloat[\label{fig:animo-edge-ui}]{\includegraphics[width=.4\textwidth]{images/animo-edge-ui}}
\caption{ANIMO user interface. (a) The main window of Cytoscape, with the ANIMO
plug-in showing a model network and the result of a simulation. (b) Editing the parameters for an ANIMO interaction.}
\end{center}
\end{figure}

@@

The scenarios are used to choose which are the nodes to be taken into account
when computing the speed at which an interaction occurs.
\begin{description}
  \item[Scenario 1] the simplest of the three, this scenario approximates an interaction
    without taking into account the abundance of substrate: the activation/inhibition rate
    depends only on the activity level of the upstream node. In the MEK $\rightarrow$ ERK
    example, the rate of the interaction would be linearly dependent on the current activity level of MEK.
  \item[Scenario 2] scenario 2 models the interaction rate
    to be directly dependent both on the activity
    level of the upstream node and on the availability of substrate. By substrate we
    mean the inactive downstream node in case of activation interaction, and active downstream node in case of inhibition.
    In the MEK $\rightarrow$ ERK example, the interaction rate is linearly dependent on
    both the activity level of MEK and the inactive fraction of ERK.
    The inactive level of a node is computed by subtracting the current activity level from the maximum activity.
  \item[Scenario 3] the nodes on which this scenario depends can be chosen directly by the user,
    and can be the active or inactive fraction of any two nodes in the network.
    This scenario can be used to represent an AND-gate, where two nodes are required to
    be simultaneously active in order to influence their target.
\end{description}

@@

In the future, we plan to improve our parameter analysis techniques, possibly using some of
the techniques already developed for testing parameter sensitivity~\cite{inverse-ta1, inverse-ta2} and
robustness~\cite{robust-ta1, robust-ta2, robust-ta3, robust-ta4} in \tas\ and hybrid systems more in general~\cite{decidability-ta-ha, imitator2}.
As already stated, parameter sensitivity is very important in biological models: if a treatment
derived from a model is based on narrow bounds for a parameter, that parameter must not exhibit significant
variety among patients. By introducing an automatic analysis for parameter sensitivity, our tool
could point out the zones of a network model where the user should concentrate more, for example by
improving the topology, or by more focused parameter sweeps.
Moreover, we also aim at applying some results from automata learning~\cite{test-based-modelling} to the generation of
\tas\ models based on experimental data: generating models that exhibit specific behaviors
implies a careful definition of their parameter space.
Finally, parametric model checking~\cite{param-mc} represents an interesting development direction:
having a tool automatically output the parameter intervals for which some
given outputs are observed is a highly valued resource for the biological researcher.
% Investigating the properties of the specific \tas\ models used in ANIMO
% could lead to important tractability results valid in our case, allowing us
% to apply methods not valid for \tas\ in general.
%FUTURE WORK!!
%parameter sensitivity: se un trattamento dipende da un parametro specifico, e' il caso che non si registri troppa varieta' per quel parametro tra i pazienti.
% per quei parametri non sensibili, si eviti di concentrare troppi sforzi. Il tool potrebbe automaticamente determinare le ``zone calde'' della rete
%anche alla luce di una possibile generazione di modelli t.a. basati su dati sperimentali, trovare modi intelligenti di generare i parametri
%il fatto che i modelli che usiamo possano identificare un sottinsieme ristretto di t.a. potrebbe essere vantaggioso per la trattabilita' dei problemi
%parametrized model checking: con che parametri e' possibile raggiungere questo stato?
%definisci un obiettivo di quello che vuoi ottenere, e lui ti dice i [range di] parametri che arrivano a quell'evoluzione
% Possibile future work: vedere meglio altri modi di applicare parameter setting nel contesto dei t.a., usare
% calcolo distribuito per accelerare la ricerca dei parametri

>>>>

A basic type of analysis that can be performed with ANIMO is the generation of
a simulation run, which is presented to the user in the form of a graph (to the right). The activity
graphs generated by ANIMO show the variation in activity of selected nodes over the course of the simulation run.
In addition to that, a slider placed under each graph allows the user to color the nodes in the Network panel
depending on their activity level at any point during the course of the simulation: the Legend
panel on the left links colors to activity. ANIMO models are based on the formalism of \tas~\cite{timed-automata-alur-dill},
which is not directly shown to the user. Indeed, ANIMO was designed with the aim of enabling expert biologists
to formalize their knowledge without the need for additional mathematical training.
For the details on the Timed Automata model used in ANIMO, we refer the interested reader to~\cite{animo-ieee}.

\begin{figure}[htbp]
\begin{center}
\subfloat[\label{fig:animo-ui}]{\includegraphics[width=.55\textwidth]{images/animo-ui}} \quad
\subfloat[\label{fig:animo-edge-ui}]{\includegraphics[width=.4\textwidth]{images/animo-edge-ui}}
\caption{ANIMO user interface. (a) The main window of Cytoscape, with the ANIMO
plug-in showing a network model and the result of a simulation. (b) Editing the parameters for an ANIMO interaction.}
\end{center}
\end{figure}

@@

The scenarios are used to choose which are the nodes to be taken into account
when computing the speed at which an interaction occurs.
\begin{description}
  \item[Scenario 1] the simplest of the three, this scenario approximates an interaction
    without taking into account the abundance of substrate (what is being
    changed by the interaction): the activation/inhibition rate
    depends only on the activity level of the upstream node. In the MEK $\rightarrow$ ERK
    example, the rate of the interaction would be linearly dependent on the current activity level of MEK.
  \item[Scenario 2] scenario 2 models the interaction rate
    to be directly dependent both on the activity
    level of the upstream node and on the availability of substrate. By substrate we
    mean the inactive downstream node in case of activation interaction, and active downstream node in case of inhibition.
    In the MEK $\rightarrow$ ERK example, the interaction rate is linearly dependent on
    both the activity level of MEK and the inactive fraction of ERK.
    The inactive level of a node is computed by subtracting the current activity level from the maximum activity.
  \item[Scenario 3] the nodes on which this scenario depends can be chosen directly by the user,
    and can be the active or inactive fraction of any two nodes in the network.
    This scenario can be used to represent an AND-gate, where two nodes are required to
    be simultaneously active in order to influence their target.
\end{description}

@@

In the future, we plan to improve our parameter analysis techniques, possibly using some of
the techniques already developed for testing parameter sensitivity~\cite{inverse-ta1, inverse-ta2} and
robustness~\cite{robust-ta1, robust-ta2, robust-ta3, robust-ta4} in \tas\ and hybrid systems more in general~\cite{decidability-ta-ha, imitator2}.
Techniques adopted in other formalisms such as e.g. differential equations~\cite{copasi, breach, kinfer}, rule-based languages~\cite{biocham, rulebender}
or Bayesian networks~\cite{bayes-infer} are also being
considered as a source of inspiration.
As already stated, parameter sensitivity is very important in biological models: if a treatment
derived from a model is based on narrow bounds for a parameter, that parameter must not exhibit significant
variety among patients. By introducing an automatic analysis for parameter sensitivity, our tool
could point out the zones of a network model where the user should concentrate more, for example by
improving the topology, or by more focused parameter sweeps.
Moreover, we also aim at applying some results from automata learning~\cite{test-based-modelling} to the generation of
\tas\ models based on experimental data: generating models that exhibit specific behaviors
implies a careful definition of their parameter space.
Finally, parametric model checking~\cite{param-mc} represents an interesting development direction:
having a tool automatically output the parameter intervals for which some
given outputs are observed is a highly valued resource for the biological researcher.
% Investigating the properties of the specific \tas\ models used in ANIMO
% could lead to important tractability results valid in our case, allowing us
% to apply methods not valid for \tas\ in general.
%FUTURE WORK!!
%parameter sensitivity: se un trattamento dipende da un parametro specifico, e' il caso che non si registri troppa varieta' per quel parametro tra i pazienti.
% per quei parametri non sensibili, si eviti di concentrare troppi sforzi. Il tool potrebbe automaticamente determinare le ``zone calde'' della rete
%anche alla luce di una possibile generazione di modelli t.a. basati su dati sperimentali, trovare modi intelligenti di generare i parametri
%il fatto che i modelli che usiamo possano identificare un sottinsieme ristretto di t.a. potrebbe essere vantaggioso per la trattabilita' dei problemi
%parametrized model checking: con che parametri e' possibile raggiungere questo stato?
%definisci un obiettivo di quello che vuoi ottenere, e lui ti dice i [range di] parametri che arrivano a quell'evoluzione
% Possibile future work: vedere meglio altri modi di applicare parameter setting nel contesto dei t.a., usare
% calcolo distribuito per accelerare la ricerca dei parametri
